### 2016-05-12 text/Chapter03.tex
Edits in implementation, consistency (Chapter 7)

\item We describe a number of simple calculi for tracking a wide range of contextual properties.
  The systems are adapted from diverse sources (type systems, static analyses, logics) and apply to
  various domains (cross-compilation, liveness, distributed computing, data-flow, security), but
  share a common structure.

@@

\item In addition, the coeffect systems for tracking the number of accessed past values in
  data-flow languages (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
  presents novel results and can be used to optimize data-flow programs.
\end{itemize}

@@

As mentioned in the introduction, coeffect systems in this chapter may appear as a disconnected
set of examples at first. Indeed, this section covers a diverse set of calculi including
Haskell's implicit parameters (Section~\ref{sec:applications-flat-impl}), distributed computing and
cross-compilation (Section~\ref{sec:applications-flat-distr}), liveness analysis
(Section~\ref{sec:applications-flat-live}) and data-flow (Section~\ref{sec:applications-flat-dataflow}).

@@

\subsection{Data-flow languages}
\label{sec:applications-flat-dataflow}

@@

The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an
operation that accesses a context. In case of arrays, the context is neighbourhood of a current
location in the array specified by a cursor. In this section, we make the example more concrete,
using a simpler and better studied programming model, data-flow languages.

Lucid \cite{app-lucid} is a declarative data-flow language designed by Wadge and Ashcroft. In Lucid,
variables represent streams and programs are written as transformations over streams. A function
application $\ident{square}(x)$ represents a stream of squares calculated from the stream of values $x$.

The data-flow approach has been successfully used in domains such as development of real-time embedded
application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the data-flow
paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
and implements a program to count the number of edges on a Boolean stream:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{edge} = \ident{false}~\kvd{fby}~(\ident{input}~\&\&~\ident{not}~(\kvd{prev}~\ident{input}))
\\[0.5em]
\kvd{let}~\ident{edgeCount} = \\[-0.25em]
\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~1 + (\kvd{prev}~\ident{edgeCount})\\[-0.25em]
\quad\quad\quad~~~\, \kvd{else}~\kvd{prev}~\ident{edgeCount} ~)
\end{array}
\end{equation*}
%
The construct $\kvd{prev}~x$ returns a stream consisting of previous values of the stream
$x$. The second value of $\kvd{prev}~x$ is first value of $x$ (and the first
value is undefined). The construct $y~\kvd{fby}~x$ returns a stream whose first element is the
first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
such as \ident{false} and $0$ are constant streams.

Formally, the constructs are defined as follows (writing $x_n$ for $n$-th element of a stream $x$):
%
\[
(\kvd{prev}~x)_n = \left\{
  \begin{array}{ll}
    \ident{nil}     & \; \text{if $n=0$}\\
    x_{n-1} & \; \text{if $n>0$}
  \end{array} \right.
\quad
(y~\kvd{fby}~x)_n = \left\{
  \begin{array}{ll}
    y_0     & \; \text{if $n=0$}\\
    x_n     & \; \text{if $n>0$}
  \end{array} \right.
\]
%
When reading data-flow programs, we do not need to think about variables in terms of streams --
we can see them as simple values. Most of the operations perform calculation just on the
\emph{current} value of the stream. However, the operation \kvd{fby} and \kvd{prev} are different.
They require additional \emph{context} which provides past values of variables
(for \kvd{prev}) and information about the current location in the stream (for \kvd{fby}).

The semantics of Lucid-like languages can be captured using a number of mathematical
structures. Wadge \cite{app-lucid-monads} originally defined a monadic semantics, while Uustalu
and Vene later used comonads \cite{app-dataflow-essence}. In Chapter~\ref{ch:flat}, we extend
the latter approach. The present chapter presents a sketch of a concrete data-flow semantics
defined directly on streams.

In the introductory example with array access patterns, we used coeffects to track the range
of values accessed. In this section, we look at a simpler example -- we only consider the
\kvd{prev} operation and track the maximal number of \emph{past values} needed. This is an
important information for efficient implementation of data-flow languages. When we can guarantee
that at most $x$ past values are accessed, the values can be stored in a pre-allocated buffer
rather than using \eg~on-demand computed lazy streams.

@@

\figcaption{Coeffect rules for tracking context-usage in data-flow language}
\label{fig:applications-flat-dataflow}
\end{figure}

@@

The (\emph{app}) rule follows the same intuition as for liveness. It combines \emph{sequential}
and \emph{pointwise} composition of semantic functions. In case of data-flow, the operations
combine annotations using $+$ and \emph{max} operations:
%
\begin{equation*}
\begin{array}{lll}
f : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 \qquad\qquad&
g : \tau_2 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
g \circ f : \tau_1 \xrightarrow{\cclrd{m+n}} \tau_3 \qquad\qquad
\\
f : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 \qquad\qquad&
h : \tau_1 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{\textnormal{max}}(\cclrd{m}, \cclrd{s})} (\tau_2 \times \tau_3) \qquad
\end{array}
\end{equation*}
%
Sequential composition adds the annotations. The function $f$ needs $\cclrd{m}$ past values to
produce a single $\tau_2$ value. To produce two $\tau_2$ values, we thus need $\cclrd{m}+1$ past
values of $\tau_1$; to produce three $\tau_2$ values, we need $\cclrd{m}+2$ past values of $\tau_1$,
and so on. To produce $\ident{n}$ past values that are required as the input of $g$, we need
$\cclrd{m+n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
the same stream to evaluate functions requiring $\cclrd{m}$ and $\cclrd{n}$ past values, and so it
needs maximum of the two at most.

@@

\paragraph{Semantics.}
The language discussed in this section is a \emph{causal} data-flow language. This means that
a computation can access \emph{past} values of the stream (but not future values). In the semantics,
we again need richer structure over the input.

Uustalu and Vene \cite{comonads-notions} model causal data-flow computations using a non-empty list
$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$ over the input. A function $\tau_1 \rightarrow \tau_2$
is thus modelled as $\ident{NeList}~\tau_1 \rightarrow \tau_2$. This model is difficult to implement
efficiently, as it creates unbounded lists of past elements.

@@

\figcaption{Semantics showing how past values are accessed in a data-flow language}
\label{fig:applications-flat-dfsem}
\end{figure}

@@

\paragraph{Summary.}
Type systems have been used in the context of data-flow languages, for example to check
initialization properties \cite{app-dataflow-init}, but to our knowledge, not for checking
the maximal number of required past values. Thus this section serves not just as an example,
but also shows how coeffects can lead to novel results.

The most interesting point about the data-flow system is that it is remarkably similar to our
earlier liveness example. In the type system, abstraction (\emph{abs}) duplicates the context
requirements and application (\emph{abs}) arises from sequential and pointwise composition.
We capture this striking similarity in Chapter~\ref{ch:flat}. Before doing that, we look at one
more example and then explore the \emph{structural} class of systems.

@@

\subsection{Permissions and safe locking}
In the implicit parameters and data-flow examples, the context provides additional resources or
values that may be accessed at runtime. However, coeffects can also track \emph{permissions} or
\emph{capabilities} to perform some operation. We can invert the intuition behind liveness and
use it as a trivial example. When the context is live, it contains a \emph{permission} to access
variables. In this section, we briefly consider a system for safe locking of Flanagan and
Abadi~\cite{app-safe-locking} as one, more advanced example. Calculus of capabilities of
Cray et al.~\cite{app-capabilities} is discussed later in Section~\ref{sec:applications-active-ccc}.

@@

In this section, we look at three examples in detail -- we revisit liveness and show a practically
useful structural version of the system; we consider an example inspired by linear logic; finally,
we revisit data-flow to get a more precise analysis. Although quite different, the common pattern
among these three examples is somewhat easier to see, because they all track information about
variable usage. We finish the section with a brief outline of several other applications.

@@

\subsection{Data-flow languages revisited}
\label{sec:applications-structural-dataflow}

When discussing data-flow languages in an earlier section, we said that the context provides
past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as
a \emph{flat} property, which gives us a system that keeps the same number of past values for
all variables. However, data-flow can also be adapted to a structural system which keeps the number
of required past values individually for each variable. Consider the
following example:
%
\begin{equation*}
\kvd{let}~\ident{offsetAdd} = \ident{left} + \kvd{prev}~\ident{right}
\end{equation*}
%
The value \ident{offsetAdd} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
requires $1$ past values of both streams in the variable context.

Turning a flat data-flow system to a structural data-flow system is a change similar to the one
between flat ans structural liveness. In case of liveness analysis, we included the flat system
only as an illustration (it is not practically useful). For data-flow, the flat system is less
precise, but still practically useful (simplicity may outweigh precision).

@@

In data-flow, we annotate both used variables (\emph{var}) and unused variables (\emph{weak}) with
$0$, meaning that no past values are required. This is the same as in flat data-flow, but different
from bounded reuse and liveness (where unused variables have a different coeffect). Primitive
requirements are introduced by the (\emph{prev}) rule, which increments the annotations of
all variables.

In flat data-flow, we identified sequential composition and pointwise composition as two primitive
operations that were used in the (flat) application. In the structural system, these are used in
(\emph{app}) and (\emph{contr}). Thus application combines coeffect annotations using $+$ and
contraction using \emph{max}. This contrasts with bounded reuse, which uses $\ast$ and $+$,
respectively.

@@

\paragraph{Semantics.}
To define the semantics of our structural data-flow language, we can use the same approach as when
adapting flat liveness to structural liveness. Rather than wrapping the whole context in a type
constructor (list or option), we now wrap the individual components of the product representing
the variables in the context.

@@

\paragraph{Summary.}
Two of the structural examples shown so far (liveness and data-flow) extend an earlier flat
version of a similar system. We discuss this relation in general later. However, a flat system
can generally be turned into a structural one -- although this only gives a useful system when
the flat version captures statically scoped properties, \ie~related to variables.

The data-flow example demonstrates that the a flat system can also be turned into structural
system. In general, this only works for systems where lambda abstraction duplicates context
requirements (as in Figure~\ref{fig:applications-flat-liveness}).

@@

There is a number of systems that also capture contextual properties, but that make it possible to
\emph{change} the context -- not just by evaluating certain code block in a locally modified context
(\eg by wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example,
acquires new capabilities and returns those to the caller. Such actions appear to be closer to
effects than to coeffects. While this thesis focuses on systems with passive contexts, we briefly
consider the most important examples of the \emph{active} variant.

@@

This chapter served two purposes. The first aim was to present existing work on programming
languages and systems that include some notion of \emph{context}. Because there was no well-known
abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
principled approaches based on comonads and modal S4, ad-hoc type system extensions and static analyses
as well as approaches based on monads. We looked at a number of applications including Haskell's implicit
parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of
security properties.

>>>>

\item We describe a number of simple calculi for tracking a wide range of contextual properties.
  The systems are adapted from diverse sources (type systems, static analyses, logics) and apply to
  various domains (cross-compilation, liveness, distributed computing, dataflow, security), but
  share a common structure.

@@

\item In addition, the coeffect systems for tracking the number of accessed past values in
  dataflow languages (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
  presents novel results and can be used to optimize dataflow programs.
\end{itemize}

@@

As mentioned in the introduction, coeffect systems in this chapter may appear as a disconnected
set of examples at first. Indeed, this section covers a diverse set of calculi including
Haskell's implicit parameters (Section~\ref{sec:applications-flat-impl}), distributed computing and
cross-compilation (Section~\ref{sec:applications-flat-distr}), liveness analysis
(Section~\ref{sec:applications-flat-live}) and dataflow (Section~\ref{sec:applications-flat-dataflow}).

@@

\subsection{Dataflow languages}
\label{sec:applications-flat-dataflow}

@@

The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an
operation that accesses a context. In case of arrays, the context is neighbourhood of a current
location in the array specified by a cursor. In this section, we make the example more concrete,
using a simpler and better studied programming model, dataflow languages.

Lucid \cite{app-lucid} is a declarative dataflow language designed by Wadge and Ashcroft. In Lucid,
variables represent streams and programs are written as transformations over streams. A function
application $\ident{square}(x)$ represents a stream of squares calculated from the stream of values $x$.

The dataflow approach has been successfully used in domains such as development of real-time embedded
application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the dataflow
paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
and implements a program to count the number of edges on a Boolean stream:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{edge} = \ident{false}~\kvd{fby}~(\ident{input}~\&\&~\ident{not}~(\kvd{prev}~\ident{input}))
\\[0.5em]
\kvd{let}~\ident{edgeCount} = \\[-0.25em]
\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~1 + (\kvd{prev}~\ident{edgeCount})\\[-0.25em]
\quad\quad\quad~~~\, \kvd{else}~\kvd{prev}~\ident{edgeCount} ~)
\end{array}
\end{equation*}
%
The construct $\kvd{prev}~x$ returns a stream consisting of previous values of the stream
$x$. The second value of $\kvd{prev}~x$ is first value of $x$ (and the first
value is undefined). The construct $y~\kvd{fby}~x$ returns a stream whose first element is the
first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
such as \ident{false} and $0$ are constant streams.

Formally, the constructs are defined as follows (writing $x_n$ for $n$-th element of a stream $x$):
%
\[
(\kvd{prev}~x)_n = \left\{
  \begin{array}{ll}
    \ident{nil}     & \; \text{if $n=0$}\\
    x_{n-1} & \; \text{if $n>0$}
  \end{array} \right.
\quad
(y~\kvd{fby}~x)_n = \left\{
  \begin{array}{ll}
    y_0     & \; \text{if $n=0$}\\
    x_n     & \; \text{if $n>0$}
  \end{array} \right.
\]
%
When reading dataflow programs, we do not need to think about variables in terms of streams --
we can see them as simple values. Most of the operations perform calculation just on the
\emph{current} value of the stream. However, the operation \kvd{fby} and \kvd{prev} are different.
They require additional \emph{context} which provides past values of variables
(for \kvd{prev}) and information about the current location in the stream (for \kvd{fby}).

The semantics of Lucid-like languages can be captured using a number of mathematical
structures. Wadge \cite{app-lucid-monads} originally defined a monadic semantics, while Uustalu
and Vene later used comonads \cite{app-dataflow-essence}. In Chapter~\ref{ch:flat}, we extend
the latter approach. The present chapter presents a sketch of a concrete dataflow semantics
defined directly on streams.

In the introductory example with array access patterns, we used coeffects to track the range
of values accessed. In this section, we look at a simpler example -- we only consider the
\kvd{prev} operation and track the maximal number of \emph{past values} needed. This is an
important information for efficient implementation of dataflow languages. When we can guarantee
that at most $x$ past values are accessed, the values can be stored in a pre-allocated buffer
rather than using \eg~on-demand computed lazy streams.

@@

\figcaption{Coeffect rules for tracking context-usage in dataflow language}
\label{fig:applications-flat-dataflow}
\end{figure}

@@

The (\emph{app}) rule follows the same intuition as for liveness. It combines \emph{sequential}
and \emph{pointwise} composition of semantic functions. In case of dataflow, the operations
combine annotations using $+$ and \emph{max} operations:
%
\begin{equation*}
\begin{array}{lll}
f : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 \qquad\qquad&
g : \tau_2 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
g \circ f : \tau_1 \xrightarrow{\cclrd{m+n}} \tau_3 \qquad\qquad
\\
f : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 \qquad\qquad&
h : \tau_1 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{\textnormal{max}}(\cclrd{m}, \cclrd{s})} (\tau_2 \times \tau_3) \qquad
\end{array}
\end{equation*}
%
Sequential composition adds the annotations. The function $f$ needs $\cclrd{m}$ past values to
produce a single $\tau_2$ value. To produce two $\tau_2$ values, we thus need $\cclrd{m}+1$ past
values of $\tau_1$; to produce three $\tau_2$ values, we need $\cclrd{m}+2$ past values of $\tau_1$,
and so on. To produce $\ident{n}$ past values that are required as the input of $g$, we need
$\cclrd{m+n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
the same stream to evaluate functions requiring $\cclrd{m}$ and $\cclrd{n}$ past values, and so it
needs maximum of the two at most.

@@

\paragraph{Semantics.}
The language discussed in this section is a \emph{causal} dataflow language. This means that
a computation can access \emph{past} values of the stream (but not future values). In the semantics,
we again need richer structure over the input.

Uustalu and Vene \cite{comonads-notions} model causal dataflow computations using a non-empty list
$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$ over the input. A function $\tau_1 \rightarrow \tau_2$
is thus modelled as $\ident{NeList}~\tau_1 \rightarrow \tau_2$. This model is difficult to implement
efficiently, as it creates unbounded lists of past elements.

@@

\figcaption{Semantics showing how past values are accessed in a dataflow language}
\label{fig:applications-flat-dfsem}
\end{figure}

@@

\paragraph{Summary.}
Type systems have been used in the context of dataflow languages, for example to check
initialization properties \cite{app-dataflow-init}, but to our knowledge, not for checking
the maximal number of required past values. Thus this section serves not just as an example,
but also shows how coeffects can lead to novel results.

The most interesting point about the dataflow system is that it is remarkably similar to our
earlier liveness example. In the type system, abstraction (\emph{abs}) duplicates the context
requirements and application (\emph{abs}) arises from sequential and pointwise composition.
We capture this striking similarity in Chapter~\ref{ch:flat}. Before doing that, we look at one
more example and then explore the \emph{structural} class of systems.

@@

\subsection{Permissions and safe locking}
In the implicit parameters and dataflow examples, the context provides additional resources or
values that may be accessed at runtime. However, coeffects can also track \emph{permissions} or
\emph{capabilities} to perform some operation. We can invert the intuition behind liveness and
use it as a trivial example. When the context is live, it contains a \emph{permission} to access
variables. In this section, we briefly consider a system for safe locking of Flanagan and
Abadi~\cite{app-safe-locking} as one, more advanced example. Calculus of capabilities of
Cray et al.~\cite{app-capabilities} is discussed later in Section~\ref{sec:applications-active-ccc}.

@@

In this section, we look at three examples in detail -- we revisit liveness and show a practically
useful structural version of the system; we consider an example inspired by linear logic; finally,
we revisit dataflow to get a more precise analysis. Although quite different, the common pattern
among these three examples is somewhat easier to see, because they all track information about
variable usage. We finish the section with a brief outline of several other applications.

@@

\subsection{Dataflow languages revisited}
\label{sec:applications-structural-dataflow}

When discussing dataflow languages in an earlier section, we said that the context provides
past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as
a \emph{flat} property, which gives us a system that keeps the same number of past values for
all variables. However, dataflow can also be adapted to a structural system which keeps the number
of required past values individually for each variable. Consider the
following example:
%
\begin{equation*}
\kvd{let}~\ident{offsetAdd} = \ident{left} + \kvd{prev}~\ident{right}
\end{equation*}
%
The value \ident{offsetAdd} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
requires $1$ past values of both streams in the variable context.

Turning a flat dataflow system to a structural dataflow system is a change similar to the one
between flat ans structural liveness. In case of liveness analysis, we included the flat system
only as an illustration (it is not practically useful). For dataflow, the flat system is less
precise, but still practically useful (simplicity may outweigh precision).

@@

In dataflow, we annotate both used variables (\emph{var}) and unused variables (\emph{weak}) with
$0$, meaning that no past values are required. This is the same as in flat dataflow, but different
from bounded reuse and liveness (where unused variables have a different coeffect). Primitive
requirements are introduced by the (\emph{prev}) rule, which increments the annotations of
all variables.

In flat dataflow, we identified sequential composition and pointwise composition as two primitive
operations that were used in the (flat) application. In the structural system, these are used in
(\emph{app}) and (\emph{contr}). Thus application combines coeffect annotations using $+$ and
contraction using \emph{max}. This contrasts with bounded reuse, which uses $\ast$ and $+$,
respectively.

@@

\paragraph{Semantics.}
To define the semantics of our structural dataflow language, we can use the same approach as when
adapting flat liveness to structural liveness. Rather than wrapping the whole context in a type
constructor (list or option), we now wrap the individual components of the product representing
the variables in the context.

@@

\paragraph{Summary.}
Two of the structural examples shown so far (liveness and dataflow) extend an earlier flat
version of a similar system. We discuss this relation in general later. However, a flat system
can generally be turned into a structural one -- although this only gives a useful system when
the flat version captures statically scoped properties, \ie~related to variables.

The dataflow example demonstrates that the a flat system can also be turned into structural
system. In general, this only works for systems where lambda abstraction duplicates context
requirements (as in Figure~\ref{fig:applications-flat-liveness}).

@@

There is a number of systems that also capture contextual properties, but that make it possible to
\emph{change} the context -- not just by evaluating certain code block in a locally modified context
(\eg by wrapping it in $\ident{prev}$ in dataflow), but also by calling a function that, for example,
acquires new capabilities and returns those to the caller. Such actions appear to be closer to
effects than to coeffects. While this thesis focuses on systems with passive contexts, we briefly
consider the most important examples of the \emph{active} variant.

@@

This chapter served two purposes. The first aim was to present existing work on programming
languages and systems that include some notion of \emph{context}. Because there was no well-known
abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
principled approaches based on comonads and modal S4, ad-hoc type system extensions and static analyses
as well as approaches based on monads. We looked at a number of applications including Haskell's implicit
parameters and type classes, dataflow languages such as Lucid, liveness analysis and also a number of
security properties.

### 2016-05-12 text/Chapter04.tex
Edits in implementation, consistency (Chapter 7)

\noindent
The liveness example is interesting because it does not require any additional syntactic extensions
to the langauge. It annotates constants and variables with $\ident{D}$ and $\ident{L}$, respectively
and it captures how those annotation propagate through the remaining language constructs.

@@

A simple dataflow langauge includes an additional construct $\kvd{prev}$ for accessing the
previous value in a stream with an additional typing rule that look as follows:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \kvd{next}~e
\end{equation*}
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
As a further example that was not covered earlier, it is also possible to combine liveness analysis
and dataflow. In the above dataflow calculus, $0$ denotes that we require the current value of some
variable, but no previous values. However, for constants, we do not even need the current value.

@@

As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
differs from the rule for effect systems in that it does not delay all context demands.
In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
can be satisfied either by the call-site or by the declaration-site. In case of dataflow and
liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
because it allows multiple valid typing derivations.

@@

\noindent
There are two possible typings allowed by the typing rules discussed in Section~\ref{sec:flat-calculus-types}
that lead to two possible meanings of the program -- evaluating to $1$ and $2$, respectively:
%
\begin{itemize}
  \item $f : \ident{num} \xrightarrow{\cclrd{\emptyset}} \ident{num}$ -- in this case, the value
    of \ident{?x} is captured from the declaration-site and the program produces $1$.
  \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
    parameter \ident{?x} is required from the call-site and the program produces $2$.
\end{itemize}
%
The coeffect calculus intentionally allows both of the options, acknowledging the fact that the
choice needs to be made for each individual concrete context-aware programming language. In the
above case, one typing derivation represents dynamic binding and the other static binding, but
more subtleties arise when the nested expression uses multiple implicit parameters.

@@

For implicit parameters we choose to follow the behaviour implemented by Haskell~\cite{app-implicit-parameters}
where function abstraction captures all parameters that are statically available at the declaration-site
and places all other demands on the call-site. For the example above, this means
that the body of $f$ captures the value of \ident{?p} available from the declaration-site
and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The
program thus evaluates to a numerical value $42$.

@@

The most interesting rule is (\emph{abs}). The body of a function requires implicit parameters
tracked by $\cclrd{r}$ and the parameters currently in (static) scope are $\cclrd{\Delta}$.
The coeffect on the declaration site becomes $\cclrd{\Delta}$ (capture all available parameters)
and the latent coeffect attached to the function becomes $\cclrd{r\setminus\Delta}$ (require any
remaining parameters from the call-site). Finally, in the (\emph{letpar}) rule, we add the newly
bound implicit parameter \ident{?p} to the static scope in the sub-expression $e_2$.

@@

Before discussing the syntactic properties of general coeffect calculus formally, it should be
clarified what is meant by providing a ``pathway to operational semantics'' in this section. We do
that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
Writing $\subst{e_1}{x}{e_2}$ for a standard capture-avoiding syntactic substitution, the following
equations define four syntactic reductions on the terms:
%
\begin{equation*}
\begin{array}{rllcl}
(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
\end{array}
\end{equation*}
%
The rules capture syntactic reductions that can be performed in a general calculus, without any
knowledge of the specific notion of context. If the reductions preserve the type of the expression
(type preservation), then operational semantics can be defined as a repeated application of the
rules, together with additional domain-specific rules for each context-aware langauge, until
a specified normal form (\ie~a value) is reached.

@@

An inherent property of flat coeffect systems is the ambiguity of the typing for lambda
abstraction. The body of a function requires certain context, but the context can be often
provided by either the declaration-site or the call-site. Resolving this ambiguity has to be
done differently for each concrete coeffect system, depending on its specific notion of context.
We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec:flat-unique}
and noted that the result for dataflow and liveness generalizes for any coeffect calculus
with strictly oriented coeffect algebra (Remark~\ref{thm:flat-alt-abs}).

>>>>

\noindent
The liveness example is interesting because it does not require any additional syntactic extensions
to the language. It annotates constants and variables with $\ident{D}$ and $\ident{L}$, respectively
and it captures how those annotation propagate through the remaining language constructs.

@@

A simple dataflow language includes an additional construct $\kvd{prev}$ for accessing the
previous value in a stream with an additional typing rule that look as follows:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \kvd{next}~e
\end{equation*}
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
As a further example that was not covered earlier, it is also possible to combine liveness analysis
and dataflow. In the above dataflow calculus, $0$ denotes that we require the current value of some
variable, but no previous values. However, for constants, we do not even need the current value.

@@

As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
differs from the rule for effect systems in that it does not delay all context demands.
In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
can be satisfied either by the call site or by the declaration-site. In case of dataflow and
liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
because it allows multiple valid typing derivations.

@@

\noindent
There are two possible typings allowed by the typing rules discussed in Section~\ref{sec:flat-calculus-types}
that lead to two possible meanings of the program -- evaluating to $1$ and $2$, respectively:
%
\begin{itemize}
  \item $f : \ident{num} \xrightarrow{\cclrd{\emptyset}} \ident{num}$ -- in this case, the value
    of \ident{?x} is captured from the declaration-site and the program produces $1$.
  \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
    parameter \ident{?x} is required from the call site and the program produces $2$.
\end{itemize}
%
The coeffect calculus intentionally allows both of the options, acknowledging the fact that the
choice needs to be made for each individual concrete context-aware programming language. In the
above case, one typing derivation represents dynamic binding and the other static binding, but
more subtleties arise when the nested expression uses multiple implicit parameters.

@@

For implicit parameters we choose to follow the behaviour implemented by Haskell~\cite{app-implicit-parameters}
where function abstraction captures all parameters that are statically available at the declaration-site
and places all other demands on the call site. For the example above, this means
that the body of $f$ captures the value of \ident{?p} available from the declaration-site
and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The
program thus evaluates to a numerical value $42$.

@@

The most interesting rule is (\emph{abs}). The body of a function requires implicit parameters
tracked by $\cclrd{r}$ and the parameters currently in (static) scope are $\cclrd{\Delta}$.
The coeffect on the declaration site becomes $\cclrd{\Delta}$ (capture all available parameters)
and the latent coeffect attached to the function becomes $\cclrd{r\setminus\Delta}$ (require any
remaining parameters from the call site). Finally, in the (\emph{letpar}) rule, we add the newly
bound implicit parameter \ident{?p} to the static scope in the sub-expression $e_2$.

@@

Before discussing the syntactic properties of general coeffect calculus formally, it should be
clarified what is meant by providing a ``pathway to operational semantics'' in this section. We do
that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
Writing $\subst{e_1}{x}{e_2}$ for a standard capture-avoiding syntactic substitution, the following
equations define four syntactic reductions on the terms:
%
\begin{equation*}
\begin{array}{rllcl}
(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
\end{array}
\end{equation*}
%
The rules capture syntactic reductions that can be performed in a general calculus, without any
knowledge of the specific notion of context. If the reductions preserve the type of the expression
(type preservation), then operational semantics can be defined as a repeated application of the
rules, together with additional domain-specific rules for each context-aware language, until
a specified normal form (\ie~a value) is reached.

@@

An inherent property of flat coeffect systems is the ambiguity of the typing for lambda
abstraction. The body of a function requires certain context, but the context can be often
provided by either the declaration-site or the call site. Resolving this ambiguity has to be
done differently for each concrete coeffect system, depending on its specific notion of context.
We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec:flat-unique}
and noted that the result for dataflow and liveness generalizes for any coeffect calculus
with strictly oriented coeffect algebra (Remark~\ref{thm:flat-alt-abs}).
